---
title: WSP Data Analysis
author: Part II of the Uncovering Insights series
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---



# Introduction

The is the second part of the  exercise on water supply data analysis. We start by first loading the required data sets, and preparing the data for analysis:

```{r include=FALSE }
# Setting repository
options(repos='http://cran.rstudio.com/')
# Load required packages
library(prettydoc)
library(ggplot2)
library(plotly)
library(reshape2)
library(dplyr)

# Load Data
data<- read.csv("mcpl_report_validated.csv")
data_pr<-read.csv("mcpl_profile_validated.csv")
data_pr<-data_pr[,c("town","popn_in_service_area")]
data<-left_join(data,data_pr)
```

Let us start by looking at the variables at our disposal:

```{r}
# Picking required columns
names(data)
```

## PICKING ANALYSIS VARIABLES

We will be looking at the following variables for today's analysis:

* <b>prod_m3_per_day</b>              
* <b>consumption_lpcd</b>
* <b>production_lpcd</b> 
* <b>cost_per_m3_water_used</b>
* <b>nrw</b>

## BUILDING PLOTTING FUNCTIONS

We will be building the plotting functions once again:

```{r}
chartPlot<- function (x, rows,sortvar) {
    chartData <- transform(x,town = reorder(town,rank(eval(parse(text=sortvar)))))
    chartData<-melt(chartData,id.vars='town')
    plot<-ggplot(chartData,aes(x=town, y=value)) +
            geom_bar(width=1,aes(fill = variable),stat="identity") +
            coord_flip() +
            facet_wrap("variable", scales = "free", nrow = rows) +
            scale_x_discrete(name="") +
            scale_y_continuous(name="") +
            theme(text = element_text(size=8), legend.position="none",axis.ticks.x=element_blank()) +
            guides(fill=FALSE) +
            geom_text(aes(label = round(value,1)),size = 2.5, position = position_stack(vjust = 0.5))
     
    return(plot)
}

groupedBarPlot<- function(x, sortvar) {
  chartData <- transform(x, town = reorder(town, rank(eval(parse(text=sortvar)))))
  chartData<-melt(chartData, id.vars = 'town')
  plot<-ggplot(chartData,aes(x=town, y=value)) +   
          scale_x_discrete(name="") +
          scale_y_continuous(name="") +
          geom_bar(width=0.6, aes(fill = variable), position = "dodge", stat="identity") +
          theme(text = element_text(size=8), legend.position="bottom", axis.ticks.x=element_blank(), axis.text.x =element_text(angle = 90,hjust = 1)) +
          guides(fill=FALSE)
return(plot)
}
```

## DATA ANALYSIS

```{r}
# Calcuating people per connection
chartData<-data[, c("town","production_lpcd", "consumption_lpcd", "prod_m3_per_day", "nrw","popn_covered", "total_taps", "cost_per_m3_water_used")]
chartData$people_per_conn <- chartData$popn_covered/chartData$total_taps
ggplotly(chartPlot(chartData[,c(1,3)], 1, "consumption_lpcd"))

summary(chartData$consumption_lpcd)

ggplotly(chartPlot(chartData[,c(1,2)], 1, "production_lpcd"))

chartData$extra<-chartData$production_lpcd-chartData$consumption_lpcd
ggplotly(chartPlot(chartData[,c(1,3,2,10)], 1, "extra"))
summary(chartData$extra)

chartData$elmpc<-chartData$extra*chartData$cost_per_m3_water_used
ggplotly(chartPlot(chartData[,c(1,11)], 1, "elmpc"))

chartData$tlm<-chartData$elmpc*chartData$popn_covered
ggplotly(chartPlot(chartData[,c(1,12)], 1, "tlm"))
summary(chartData$tlm)

```

