---
title: WSP Data Analysis
author: Part II of the Uncovering Insights series
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---



# Introduction

The is the second part of the  exercise on water supply data analysis. We start by first loading the required data sets, and preparing the data for analysis:

```{r include=FALSE }
# Setting repository
options(repos='http://cran.rstudio.com/')
# Load required packages
library(prettydoc)
library(ggplot2)
library(plotly)
library(reshape2)
library(dplyr)

# Load Data
data<- read.csv("./data/mcpl_report_validated.csv")
data_pr<-read.csv("./data/mcpl_profile_validated.csv")
data_pr<-data_pr[,c("town","water_consumption_sold_or_used", "operational_cost_per_m3_prodn")]
data<-left_join(data,data_pr)
```

Let us start by looking at the variables at our disposal:

```{r}
# Picking required columns
names(data)
```

## PICKING ANALYSIS VARIABLES

We will be looking at the following variables for today's analysis:

* <b>prod_m3_per_day</b>: As the name suggests, it is nothing but the average daily amount of water produced in a service area (expessed in meter cubes per day)
* <b>water_consumption_sold_or_used</b>: the average daily amount of water consumed in a service area (expessed in meter cubes per day)
* <b>operational_cost_per_m3_prodn</b>: The average operational cost for producing one meter cube of water in each service area
* <b>total_taps</b>: Total number of connections within a given service area

## BUILDING PLOTTING FUNCTIONS

We will be building the plotting functions once again:

```{r}
chartPlot<- function (x, rows,sortvar) {
    chartData <- transform(x,town = reorder(town,rank(eval(parse(text=sortvar)))))
    chartData<-melt(chartData,id.vars='town')
    plot<-ggplot(chartData,aes(x=town, y=value)) +
            geom_bar(width=1,aes(fill = variable),stat="identity") +
            coord_flip() +
            facet_wrap("variable", scales = "free", nrow = rows) +
            scale_x_discrete(name="") +
            scale_y_continuous(name="") +
            theme(text = element_text(size=8), legend.position="none",axis.ticks.x=element_blank()) +
            guides(fill=FALSE) +
            geom_text(aes(label = round(value,1)),size = 2.5, position = position_stack(vjust = 0.5))
     
    return(plot)
}

groupedBarPlot<- function(x, sortvar) {
  chartData <- transform(x, town = reorder(town, rank(eval(parse(text=sortvar)))))
  chartData<-melt(chartData, id.vars = 'town')
  plot<-ggplot(chartData,aes(x=town, y=value)) +   
          scale_x_discrete(name="") +
          scale_y_continuous(name="") +
          geom_bar(width=0.6, aes(fill = variable), position = "dodge", stat="identity") +
          theme(text = element_text(size=8), legend.position="bottom", axis.ticks.x=element_blank(), axis.text.x =element_text(angle = 90,hjust = 1)) +
          guides(fill=FALSE)
return(plot)
}
```

## DATA ANALYSIS

```{r}
# Calcuating people per connection
chartData<-data[, c("town", "prod_m3_per_day", "total_taps","popn_covered" , "water_consumption_sold_or_used", "operational_cost_per_m3_prodn" )]
chartData$people_per_conn <- chartData$popn_covered/chartData$total_taps
chartData$prod_per_day_connection<-chartData$prod_m3_per_day/chartData$total_taps
chartData$cons_per_day_connection<-chartData$water_consumption_sold_or_used/chartData$total_taps

ggplotly(chartPlot(chartData[,c(1,8,9)], 1, "cons_per_day_connection"))
```

Removing Hemja as an outlier, we have:

```{r}
ggplotly(chartPlot(chartData[-which(chartData$town=="Hemja"),c(1,8,9)], 1, "prod_per_day_connection"))
```

Calculating the difference to estimate extra water production per day per connection, and plotting the same we have:

```{r}
chartData$ewp_perday_perconn<- chartData$prod_per_day_connection-chartData$cons_per_day_connection
ggplotly(chartPlot(chartData[-which(chartData$town=="Hemja"),c(1,10)], 1, "ewp_perday_perconn"))
```

Calculating the average operational cost per m3 of production that could be saved in a day:

```{r}
chartData$pot_op_cost_sav_daily<- chartData$ewp_perday_perconn * chartData$operational_cost_per_m3_prodn
ggplotly(chartPlot(chartData[-which(chartData$town=="Hemja"),c(1,11)], 1, "pot_op_cost_sav_daily"))
```

Calculating max potential savings in a year:

```{r}
chartData$pot_op_cost_sav_yearly<- chartData$pot_op_cost_sav_daily * chartData$total_taps * 365
ggplotly(chartPlot(chartData[-which(chartData$town=="Hemja"),c(1,12)], 1, "pot_op_cost_sav_yearly"))

# Calculate summary stats:
summary(chartData$pot_op_cost_sav_yearly)
```

